Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-17 paper.tex
add citation
Add citation and update statement on coordinate transformation

\subsection{Case study: Our Future Health genotype release}

Our Future Health is the UK’s largest ever health research programme. 
It is designed to help people live healthier lives for longer through the 
discovery and testing of more effective approaches to prevention, earlier 
detection and treatment of diseases.
Our Future Health is inviting millions of people, from all backgrounds and 
from right across the UK to take part. Volunteers are providing information 
about their health and lifestyles to create an incredibly detailed picture 
that represents the whole of the UK.

% this section is probably better placed in acknowledgements?
This study makes use of de-identified data held by Our Future Health. 
We would like to acknowledge all the research participants who have donated 
their data to the Our Future Health research programme 
\url{https://a.storyblok.com/f/228028/x/f45d0da22a/our-future-health-publication-policy-version-1_4.pdf}

Release 9 (\url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9}) 
was made available in December 2024. The genetic component includes genotype data 
from 651,050 individuals on 707,522 variants. Here we use the genotype data from 
chromosome 22 to assess the performance of \texttt{vcf2zarr}. The dataset consists of
three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling 
10,221 variants.

@@

%% FIXME: reference regarding VCF POS?
The spruce data set poses an interesting challenge to the VCF storage
format due to the large chromosome sizes (>1Gbp). The VCF POS field is
encoded in 32-bit format (corresponding to a maximum of approximately
537Mbp), which mandates splitting up larger chromosomes into chunks,
adding additional complexity to downstream analyses. Zarr does not
impose this limitation, and we therefore demonstrate the versatility
of Zarr by updating contig information and the coordinate system for
the spruce dataset.

>>>>

@@

Large genomes from non-model organisms pose particular problems for
analysis workflows that often are tailored for human conditions. For
instance, the binary Sequence Alignment/Map index format BAI has a 512
Mbp limit, but spruce chromosomes range in size from 1.0 to 1.7 Gbp.
The spruce BAM files used for variant calling were BAI indexed, which
motivated the partitioning into multiple VCF files in the first place,
adding complexity to downstream processing.

A similar limitation is imposed by the VCF POS field which is encoded
in 32-bit format and can hold chromomosomes up to 2.14 Gbp. Although
spruce does not have chromosomes that exceed the limit, other species
do, such as the recently sequence mistletoe, with chromosomes larger
than 8 Gbp~\cite{dtol_mistletoe}. Zarr does not impose this
limitation, and the coordinate system could easily be converted to
true genome coordinates for storage in Zarr. This updating procedure
could be applied to any species of choice.
